Efficient Computation of Extended Sources

### 2021-05-21 repro/papers/sseprecond/paper.tex
interim

The surface source extension replaces point sources on or near a
surface in $\bR^3$ with source functions confined to the same
surface. The simplest example of this extended geometry specifies a
plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
acoustic modeling, surface sources are combinations of constitutive law
defects and loads normal to the surface, localized on $z=z_s$. That
is, right-hand sides in the system $L\bu=\bff$ take the form
$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

@@

${\cal S}^{\pm}$ is not stably invertible, in fact having rank 1 (in
an approximate sense - see the identity \ref{eqn:lamnull}). The
diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
its information, and it is in terms of these that a sensible inverse problem
is defined.

@@

Discretization is conventional, with a rectangular grid and staggered
finite difference scheme \cite[]{Virieux:84} of order 2 in time and
2$k$ in space; for most of the experiments reported below, $k=4$.
Sampling operators such as $P_r$ are implemented via linear
interpolation, and source insertion via adjoint linear interpolation
(as noted above, in the continuum limit, sources are represented via
adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
$\Delta x = 20$ m, and limit the temporal frequency of the computed
traces to 12.5 Hz. We also explore the dependence of a few results on frequency, using
$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
maintaining 8 samples per wavelength.

@@

@@

These apparently harmless relations have a profound consequence,
because of a final step: identify the boundary data $\phi$ and $\psi$,
so far arbitrary, as the traces of $p^{\pm}_0$ and $v^{\pm}_{0,z}$ on $z=z_s$,
that is, set
\begin{eqnarray}
  \phi & = & P_s p^{\pm}_0, \nonumber \\
  \psi & = & P_s v^{\pm}_{0,z}.
             \label{eqn:bv0}
\end{eqnarray}
Since the traces on $z=z_s$ of the pressure fields $p^{\pm}_0$ and $p^{\pm}_1$ are
the same, and both are pressure components of solutions of the same homogeneous
acoustic system in $z>z_s$, the standard energy identity implies that
$(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_1,\bv^{\pm}_1)$ in
$z > z_s$. Similarly, $P_sv^{\pm}_{0,z}$ and $P_sv^{\pm}_{2,z}$ are
the same, whence $(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_2,\bv^{\pm}_2)$ in
$z>z_s$.

>>>>

The surface source extension replaces point sources on or near a
surface in $\bR^3$ with source functions confined to the same
surface. The simplest example of this extended geometry specifies a
plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
acoustic modeling, surface sources are combinations of constitutive law
defects and loads normal to the surface, localized on $z=z_s$. That
is, right-hand sides in the system $L\bu=\bff$ take the form
$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} +
                                                f_s{\bf e} \delta(z-z_s),\nonumber \\
p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

@@

${\cal S}^{\pm}$ is not stably invertible (equation \ref{eqn:lamnull}). The
diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
its information, and it is in terms of these that a sensible inverse problem
is defined.

@@

Discretization is conventional, with a rectangular grid and staggered
finite difference scheme \cite[]{Vir:84} of order 2 in time and
2$k$ in space; for most of the experiments reported below, $k=4$.
Sampling operators such as $P_r$ are implemented via linear
interpolation, and source insertion via adjoint linear interpolation
(as noted above, in the continuum limit, sources are represented via
adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
$\Delta x = 20$ m, and limit the temporal frequency of the computed
traces to 12.5 Hz. We also explore the dependence of a few results on frequency, using
$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
maintaining 8 samples per wavelength. All computations are carried out
in single precision.

\cite{GeoPros:11} gives a description of the code
implementation, out-of-date in a few respects but overall accurate.

@@

The implemented code uses the discrete adjoint state method and auto-generated code \cite[]{TapenadeRef13}, to
assure that the computed adjoint operators are adjoint at the level of
machine precision to the computed operators. The reverse-time storage
issue is resolved through the optimal checkpointing technique
\cite[]{Griewank:book,Symes:06a-pub}, again without loss of precision.

Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
\begin{itemize}
\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
\end{itemize}
Ideally, the differences of these inner products should be at most a
relatively small multiple of machine precision, {\em relative} to the
products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (division by these quantities makes the result
dimensionless and scale-independent). The operator norm $\| S^+_{z_s,z_r}\|$ is
computationally inaccessible, so instead I used the smaller quantities
$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
overestimating the relative error between the inner products. In all
cases, over a very large number of random inputs, the largest observed
relative error estimate was $O(10^{-9})$, well under the appropriate limit for
single precision.

\section{Pressure-to-Source}

Since the system \ref{eqn:awepm} has a unique solution by standard
theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
prescribe both pressure and normal velocity on a surface such as
$z=z_s$. So the columns of the matrix operator
${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
section I will explain this relation; it involves the {\em
  pressure-to-source} map. This operator also turns out to be the
principle component of a preconditioning strategy for iterative
solution of the optimization problem \ref{eqn:esis}, so I will devote
some effort to its proper definition. It is closely related to the
Dirichlet-to-Neumann operator mentioned in the introduction.

While it is not possible to prescribe both pressure and velocity on
$z=z_s$ in solutions of \ref{eqn:awepm}, it is possible to
prescribe pressure only, for instance: if the function $\phi$ on
the surface $z=z_s$ satisfies suitable conditions, for
example the downgoing constraint mentioned earlier, a unique solution
exists for the acoustic system in both half-spaces $\pm z > z_s$:
\begin{eqnarray}
\label{eqn:awe0}
  \frac{1}{\kappa}\frac{\partial p_{\pm}}{\partial t} & = & - \nabla \cdot \bv_{\pm}, \nonumber \\
  \rho\frac{\partial \bv_{\pm}}{\partial t} & = & - \nabla
                                                    p_{\pm}, \nonumber \\
  p_{\pm} & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
  \bv_{\pm} & = & 0 \mbox{ for } t \ll 0, \nonumber\\
  \lim_{z \rightarrow z_s^{\pm}}p_{\pm}(x,y,t,z)& =& \phi(x,y,t).
\end{eqnarray}
Note that the subscript $\pm$ here refers to the sign of $z-z_s$, as opposed
to the superscript ${\pm}$, which refers to the sign of $t$ throughout
this paper.

From the boundary condition (last equation in \ref{eqn:awe0}), one
sees that the pressures $p_{\pm}$ in the two half-spaces have the same
limit at the boundary $z=z_s$. Stick the two half-space
solutions together to form an acoustic field $(p^+,\bv^+)$ in all of
space-time, that is,
\begin{equation}
  \label{eqn:awealt}
  p^+(x,y,z,t) =
  \left\{
    \begin{array}{c}
      p_+(x,y,z,t) \mbox{ if } z>0,\\
      p_-(x,y,z,t) \mbox{ if } z<0,
    \end{array}
  \right.
\end{equation}
and a similar definition for $\bv^+$. Then $p^+$ is continuous across
$z=z_s$, and the boundary condition in system \ref{eqn:awe0} may be
written as $P_sp^+=\phi$.

The same construction can be carried out in the anti-causal sense,
with anti-causal half-space solutions glued together to form a
full-space distribution solution $(p^-,\bv^-)$, with the property that
$p^-$ is continuous across $z=z_s$ and $P_sp^-=\phi$.

The reader may object that the notation $(p^\pm,\bv^\pm)$ is already in
use, for the solution of \ref{eqn:awepm}. This objection is
valid. However, {\em in the sense
  of distributions}, $(p^{\pm},\bv^{\pm})$ as defined in display
\ref{eqn:awealt}, is {\em exactly} the causal solution of \ref{eqn:awepm}
for the choice $h_s = -[v^{\pm}_{z}]|_{z=z_s}, f_s=0$, as follows from a
simple integration-by-parts calculation. So the notation is consistent!

The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
\begin{equation}
  \label{eqn:deflam}
  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{1,z}]|_{z=z_s}
\end{equation}
The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.

Otherwise put, $S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s} \phi = \phi$, so
$\Lambda^{\pm}_{z_s}$ is inverse to $S^{\pm}_{z_s,z_s}$. The relation
\ref{eqn:trtrcomp} implies in turn that
\begin{equation}
  \label{eqn:lamadj}
  (\Lambda^{\pm}_{z_s})^T = - \Lambda^{\mp}_{z_s}
\end{equation}

There is also a {\em velocity-to-source} operator. For the solution
$(p^{\pm},\bv^{\pm})$ of system \ref{eqn:awepm} with $h_s=0$, the
normal component of velocity, $v^{\pm}_z$, is continuous across
$z=z_s$, and the velocity source (vertical load)
$f_s=-[p^{\pm}]_{z=z_s}$. I will not name the velocity-to-source
operator, as it does not appear explicitly in the developments to
follow. As will be seen, it is essentially the inverse of the
pressure-to-source operator.

While $\Lambda^{\pm}_{z_s}$ is positive semi-definite, it is not
symmetric. However, it is {\em approximately symmetric} in the
high-frequency sense. This fact follows from a
geometric optics analysis of the half-space solution,
assuming (as always) downgoing data:
\begin{equation}
  \label{eqn:lamappsim}
  (\Lambda^{\pm}_{z_s})^T \approx \Lambda^{\pm}_{z_s}.
\end{equation}

The analysis also reveals that the solution components not continuous
at $z=z_s$ are odd there:
\begin{equation}
  \label{eqn:odd1}
  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
  v^{\pm}_{z}
\end{equation}
for the solution of \ref{eqn:awepm} with $f_s=0$.
Similarly, 
\begin{equation}
  \label{eqn:odd2}
  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
  p^{\pm}
\end{equation}
for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
means in the sense of high frequency asymptotics, that is, that the
difference between the two sides is relatively smooth, hence small if
the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
\begin{equation}
  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
  P_sv^{\pm}_{z}
  \label{eqn:tracejump10}
\end{equation}
Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
\begin{equation}
  \label{eqn:tracejump20}
  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
\end{equation}
Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
as a solution of the acoustic wave system in the half-space
$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
solution with this boundary value is also the restriction to $z>z_s$
of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
\begin{equation}
  \label{eqn:hfcondn}
  h_s =\Lambda^{\pm}_{z_s}f_s,
\end{equation}
then the pressure boundary value $P_sp^{\pm}$ is the
same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
that
\begin{equation}
  \label{eqn:snull}
  {\cal S}^{\pm}_{z_s,z_r}(-\Lambda^{\pm}_{z_s}f_s,f_s)^T \approx 0.
\end{equation}

Equation \ref{eqn:snull} states the relation between the columns of $
{\cal S}^{\pm}_{z_s,z_r}$ mentioned in the introduction to this
section.

To give an example of \ref{eqn:snull} in action, it is necessary to
create downgoing data and the action of the operator $\Lambda^{\pm}$
on a downdoing field data $f_s$ on $z=z_s$. Note that a point source
on $z=z_s$ creates high frequency energy traveling on rays parallel
and nearly parallel to $z=z_s$, so that won't do. One simple approach
is to create pressure and normal velocity data on $z=z_s$ by placing a
point source at a depth $z_d<z_s$. Since the examples used here are
homogeneous in $z<z_s$, the traces extracted from the resulting
causal pressure and velocity data on $z=z_s$ are {\em a priori}
downgoing, hence related by the operator $\Lambda^+_{z_s}$.

For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}. Since the
mechanical parameters in the homogeneous and lens models are the same
for $z<z_s$, and no rays return to this zone in either model, these
data are asymptotically the same for both models, and I show only the
homogenous medium results. Regard these gathers as the trace
$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
Equations \ref{eqn:tracejump10} and
\ref{eqn:tracejump20} show that these differ by a factor of -2 from
source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
with $h_s=0$ and $f_s=0$ respectively.

These source functions satisfy
the relation \ref{eqn:hfcondn}, therefore source vectors $(h_s,0)$ and
$(0, f_s)$ generate {\em the same acoustic field} $(p^+,\bv^+)$ in $z>z_s$. Figures
\ref{fig:drecphh0}, \ref{fig:dfwdphh0}, \ref{fig:daltphh0} (homogeneous model) and
\ref{fig:drecplh0}, \ref{fig:dfwdplh0}, \ref{fig:daltplh0} (lens model) show the pressure
gathers extracted at $z_r=1000$ m for the point source at $z=z_d$ and
for the two choices of extended source at $z=z_s$, on the same color
scale. The obvious similarity between the fields generated by the two
extended sources,
predicted by equation \ref{eqn:snull}, is confirmed by trace
comparisons in figures \ref{fig:drecphh0tr21}-\ref{fig:daltplh0tr81}.

\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

\plot{drecphh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the homogeneous model described in the 
  text, with point pressure source (constitutive defect) at $z_d=3500$
  m, $x_d=3500$ m.}

\plot{dfwdphh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the homogeneous model described in the 
  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given by $-2$
  times the field depicted in Figure \ref{fig:dsrcvzhh0}
  ($h_s=-2P_sv_z^+$) and zero velocity source (vertical load) 
  ($f_s=0$).}

\plot{daltphh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the homogeneous model described in the 
  text, with extended velocity source (vertical load) on $z=z_s=3000$ m given by $-2$
  times the field depicted in Figure \ref{fig:dsrcphh0}
  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
  ($h_s=0$).}

\plot{drecplh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the lens model described in the 
  text, with point pressure source (constitutive defect) at $z_d=3500$
  m, $x_d=3500$ m.}

\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the lens model described in the 
  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given by $-2$
  times the field depicted in Figure \ref{fig:dsrcvzhh0}
  ($h_s=-2P_sv_z^+$) and zero velocity source (vertical load) 
  ($f_s=0$).}

\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the lens model described in the 
  text, with extended velocity source (vertical load) on $z=z_s=3000$ m given by $-2$
  times the field depicted in Figure \ref{fig:dsrcphh0}
  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
  ($h_s=0$).}

\plot{drecphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$)
  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
  (red).}

\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
  (red).}
\section{Approximate Inversion}

\section{Accelerated Iterative Inversion}

@@

%%%%%%%%%%%%%%%%%%%
These apparently harmless relations have a profound consequence,
because of a final step: identify the boundary data $\phi$ and $\psi$,
so far arbitrary, as the traces of $p^{\pm}_0$ and $v^{\pm}_{0,z}$ on $z=z_s$,
that is, set
\begin{eqnarray}
  \phi & = & P_s p^{\pm}_0, \nonumber \\
  \psi & = & P_s v^{\pm}_{0,z}.
             \label{eqn:bv0}
\end{eqnarray}
Since the traces on $z=z_s$ of the pressure fields $p^{\pm}_0$ and $p^{\pm}_1$ are
the same, and both are pressure components of solutions of the same homogeneous
acoustic system in $z>z_s$, the standard energy identity implies that
$(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_1,\bv^{\pm}_1)$ in
$z > z_s$. Similarly, $P_sv^{\pm}_{0,z}$ and $P_sv^{\pm}_{2,z}$ are
the same, whence $(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_2,\bv^{\pm}_2)$ in
$z>z_s$.
